---
title: "Sex differences in blood pressure polygenic risk scores"
editor_options:
  chunk_output_type: console
output:
  html_document:
    number_sections: true
---

```{r setup}
packrat::init()

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```


**Libraries**


```{r libraries}

packrat::on()
library(tidyverse)
library(survival)   # survival analysis
library(data.table) # fread() function
library(gridExtra)  # plots to grid
library(survminer)  # ggbased visualization and extra diagnostics
library(survRM2)    # restricted mean survival times
library(visdat)     # visualization of tibble and na's
library(forestplot) # forestplot
library(kableExtra) # pretty tables

#install.packages("kableExtra_1.2.1.tar.gz", repos=NULL, type="source")
#install.packages("webshot_0.5.2.tar.gz", repos=NULL, type="source")
#install.packages("knitr_1.29.tar.gz", repos=NULL, type="source")
#install.packages("plotrix_3.8-1.tar.gz", repos=NULL, type="source")


source('/finngen/red/ht_prs/sex/Functions.R')

setwd("/home/ivm/red/ht_prs/sex")
```


# Data

**Risk scores**

Calculated from UK Biobank (UKB) for Finngen Data Freeze 5 individuals using the PRS-CS18 pipeline with default parameters.


**Endpoints**

* I9_HYPTENS: Hypertension
* I9_CHD: Major coronary heart disease event

**Copying and preprocessing data**

Endpoint and covariate files transfered and unzipped. Columns selected from endpoint file prior to import to r. There was not enough memory to upload full files to R. 


```{bash, eval=F}

#Endpoint file and covariate file copied and unzipped

cp /finngen/library-red/finngen_R5/phenotype_2.0/data/finngen_R5_V2_endpoint.gz /finngen/red/ht_prs/data/
cp /finngen/library-red/finngen_R5/phenotype_2.0/data/R5_cov_pheno_1.0.txt.gz /finngen/red/ht_prs/data/  

gzip -d /finngen/red/ht_prs/data/finngen_R5_V2_endpoint.gz
gzip -d /finngen/red/ht_prs/data/R5_cov_pheno_1.0.txt.gz

#Self written perl-script to extract columns from phenotype file
perl select_columns.pl data/finngen_R5_V2_endpoint data/finngen_R5_V2_HTCVD FINNGENID I9_HYPTENS I9_HYPTENS_AGE I9_CVD_HARD I9_CVD_HARD_AGE I9_CHD I9_CHD_AGE I9_STR I9_STR_AGE FU_END_AGE

#This could be loaded directly to R, but now loading will be faster...
perl select_columns.pl data/R5_cov_pheno_1.0.txt data/R5_cov_pheno_selected FINNGENID batch BL_YEAR cohort SEX_IMPUTED PC1 PC2 PC3 PC4 PC5 PC6 PC7 PC8 PC9 PC10

```


**Loading data**

```{r load}

endpoints <- fread("/finngen/red/ht_prs/data/finngen_R5_V2_HTCVD", sep = "\t") 

sbp_prs_f <- fread("/finngen/red/ukb-bp-prs/ukb-sbp-female.profile") %>%
  select(IID, SCORE) %>%
  rename(SBP_F_SCORE=SCORE)

sbp_prs_m <- fread("/finngen/red/ukb-bp-prs/ukb-sbp-male.profile") %>%
  select(IID, SCORE) %>%
  rename(SBP_M_SCORE=SCORE)

sbp_prs_both <- fread("/finngen/red/ukb-bp-prs/ukb-sbp-both.profile") %>%
  select(IID, SCORE) %>%
  rename(SBP_SCORE_ALL=SCORE)

covs <- fread("/finngen/red/ht_prs/data/R5_cov_pheno_selected")

```


**Combining data**


```{r combine}
df <- endpoints %>%
  left_join(sbp_prs_f,    by = c("FINNGENID" = "IID")) %>%
  left_join(sbp_prs_m,    by = c("FINNGENID" = "IID")) %>%
  left_join(sbp_prs_both, by = c("FINNGENID" = "IID")) %>%
  left_join(covs, by = c("FINNGENID" = "FINNGENID")) 
 
#Removing cases without genetic information and some other preprocessing
#...and creating combined PRS score

df <- df %>%
  filter(!is.na(batch)) %>%  
  mutate_at(c("batch","cohort"),as.factor) %>%
  mutate(SBP_SCORE = if_else(as.logical(SEX_IMPUTED),SBP_F_SCORE, SBP_M_SCORE)) #%>%
 # mutate(SBP_SCALED_ALL = scale(SBP_SCORE_ALL))

dim(df)

```


**Dividing to late and early hypertension**


```{r early vs late}

df <- df %>%  
  mutate(HT_EARLY = if_else(I9_HYPTENS_AGE >= 55, as.integer(0), I9_HYPTENS),
         HT_LATE  = if_else(I9_HYPTENS_AGE <  55,  as.integer(0), I9_HYPTENS),
         AGE_CAT  = if_else(I9_HYPTENS_AGE <  55, 1,2))

df <- df %>%  
  mutate(CHD_EARLY = if_else(I9_CHD_AGE >= 55, as.integer(0), I9_CHD),
         CHD_LATE  = if_else(I9_CHD_AGE <  55,  as.integer(0), I9_CHD))

```


**Separate dataset by gender**

```{r by gender}
df_m <-  df %>%   filter(SEX_IMPUTED==0)
df_f <-  df %>%   filter(SEX_IMPUTED==1)

```


**Categorical variables & sex**

Data is categorized and scaled separately for both genders.

```{r scaling and categorizing, R.options = list(width = 120) }

#Let's add parametric versions of the scores
my_probs = c(0,0.025,0.2,0.8,0.975,1)
my_catergories =  c("<2.5%","2.5-20%","20-80%","80-97.5%",">97.5%")


#df_f & female scores
df_f <- df_f %>%  
  mutate(
    SBP_SCALED = scale(SBP_SCORE)[,1], 
    SBP_CAT = cut(SBP_SCORE, breaks=quantile(df_f$SBP_F_SCORE, probs = my_probs), 
                         labels=my_catergories, include.lowest = T))

#df_m & male scores
df_m <- df_m %>%  
  mutate(
    SBP_SCALED = scale(SBP_SCORE)[,1], 
    SBP_CAT = cut(SBP_SCORE, breaks=quantile(df_m$SBP_M_SCORE, probs = my_probs), 
                    labels=my_catergories, include.lowest = T))



##Here we add the new separately calculated columns also to original dataframe.
df <- bind_rows(df_f, df_m) %>%
  select(FINNGENID, SBP_SCALED, SBP_CAT) %>%
  right_join(df, by="FINNGENID")


dim(df)
dim(df_f)
dim(df_m)

```


#Let's create new dataframe for tables with 20-80% as reference. Needed for some tables

```{r new reference}


df_fr <- df_f %>%
  mutate(SBP_CAT = relevel(SBP_CAT, ref = "20-80%"))

df_mr <- df_m %>%
  mutate(SBP_CAT = relevel(SBP_CAT, ref = "20-80%"))

df_r <- df %>%
  mutate(SBP_CAT = relevel(SBP_CAT, ref = "20-80%"))


```


**Properties of variables **


```{r descriptive, R.options = list(width = 120) }

#Number individuals, females and males, fraction of hypertensive patients

dim(df)

df$SEX_IMPUTED %>%
  mean()

df %>% 
  group_by(SEX_IMPUTED) %>%
  summarize(n=n(), fract_HT = mean(I9_HYPTENS, na.rm=T))

sum(df_f$I9_HYPTENS, na.rm=T)
sum(df_m$I9_HYPTENS, na.rm=T)


#Average ages for HT in each riskgroup

df %>%
  mutate(HYPTENS_AGE_NA_F = if_else(I9_HYPTENS == 1 & SEX_IMPUTED == 1, I9_HYPTENS_AGE, NA_real_),
         HYPTENS_AGE_NA_M = if_else(I9_HYPTENS == 1 & SEX_IMPUTED == 0, I9_HYPTENS_AGE, NA_real_)) %>%
  group_by(SBP_CAT)%>%
  summarise_at(c("HYPTENS_AGE_NA_F","HYPTENS_AGE_NA_M"), mean, na.rm=T) %>% 
    rename(HT_AGE_F = HYPTENS_AGE_NA_F, HT_AGE_M = HYPTENS_AGE_NA_M) %>%
  ungroup() %>%
  kable(digits = 1) %>% kable_styling(full_width = F)



#Age at the end of followup

summary(df$FU_END_AGE)
sd(df$FU_END_AGE)


```    



<details><summary>**Visualisations on data**</summary>


```{r descriptive2, R.options = list(width = 120) }

#5000 random observations
df %>% 
  sample_n(5000) %>%
  vis_dat()

#Risk scores

df_tmp_f <- df_f %>% select(I9_HYPTENS_AGE, I9_HYPTENS, SBP_SCALED, SBP_CAT)
summary(df_tmp_f)


df_tmp_m <- df_m %>% select(I9_HYPTENS_AGE, I9_HYPTENS, SBP_SCALED, SBP_CAT)
summary(df_tmp_m)

par(mfrow=c(1,2))
hist(df_f$SBP_SCORE)
hist(df_m$SBP_SCORE)
par(mfrow=c(1,1))

par(mfrow=c(1,2))
hist(df_f$SBP_SCALED)
hist(df_m$SBP_SCALED)

par(mfrow=c(1,1))

#Visualisation of endpoints

ggplot(df, aes(x=I9_HYPTENS_AGE, color=as.factor(I9_HYPTENS), fill=as.factor(I9_HYPTENS))) +
    geom_histogram(alpha=0.4) +
    theme(legend.title = element_blank()) +
    facet_wrap(~SEX_IMPUTED, ncol=2)


```

</details>
<br>



# BP PRSs and hypertension


## Proportional hazard assumption: log log curve


**Model by survfit**


```{r km survfit}

km.sbp.f <- survfit(Surv(I9_HYPTENS_AGE, I9_HYPTENS) ~ SBP_CAT, data=df_f)
#print(km.sbp)
km.sbp.m <- survfit(Surv(I9_HYPTENS_AGE, I9_HYPTENS) ~ SBP_CAT, data=df_m)
#print(km.sbp)

```


**KM plot**


```{r km plot, fig.width=10, fig.height=4}


p.f <- my_ggkmplot(km.sbp.f, "HT vs. SBP score")
p.m <- my_ggkmplot(km.sbp.m, "HT vs. DBP score")

splots <- arrange_ggsurvplots(list(p.f, p.m), 
                            nrow = 1, ncol = 2, 
                            title = "Kaplan Meier curve")

ggsave(file = "km_ht_prs.png", plot = splots, height = 4, width = 10, dpi = 150)


```


**log log curve**


```{r log log, fig.width=10, fig.height=5}

cols= c("skyblue", "blue", "black", "red", "salmon")
labels= c("<2.5%","2.5-20%","20-80%","80-97.5%",">97.5%")

par(mfrow=c(1,2))

plot(km.sbp.f, fun="cloglog", xlim=c(20,100), main="SBP female: log log", xlab="Age", col = cols)
plot(km.sbp.m, fun="cloglog", xlim=c(20,100), main="SBP male: log log", xlab="Age", col = cols)

par(mfrow=c(1,1))

```

These look parallel enough. We can use cox model.


## Cox model: covariate adjusted, continuous score



**Running models**


```{r model continuous score}
#Fitting cox for females
cx.sbp.cs.f <- coxph(Surv(I9_HYPTENS_AGE, I9_HYPTENS) ~ SBP_SCALED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_f)

#Fitting cox for males
cx.sbp.cs.m <- coxph(Surv(I9_HYPTENS_AGE, I9_HYPTENS) ~ SBP_SCALED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_m)

#Both
#cx.sbp.cs.b <- coxph(Surv(I9_HYPTENS_AGE, I9_HYPTENS) ~ SBP_SCALED_ALL + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df)

```


**Table: Hazard rations:**


```{r HRs continuous score}

my_hr_table_cont2(list(cx.sbp.cs.f, cx.sbp.cs.m, cx.sbp.cs.b), c("SBP female", "SBP male", "SBP both")) %>%
  kable() %>%
  kable_styling(full_width =  F)

```



<details><summary>**All coefficients**</summary>


```{r all coefs continuous}

summary(cx.sbp.cs.f)
summary(cx.sbp.cs.m)
summary(cx.sbp.cs.b)

```


</details>
<br>




## Cox model: adjusted, continuous, by age

**Running the model:**

*Female*

```{r female continuous by age}
#Young
cx.sbp.cs.y.f <- coxph(Surv(I9_HYPTENS_AGE, HT_EARLY) 
                       ~ SBP_SCALED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_f)

#Old
cx.sbp.cs.o.f <- coxph(Surv(I9_HYPTENS_AGE, HT_LATE) 
                       ~ SBP_SCALED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_f)

```

*Male*

```{r male continuous by age}
#Young
cx.sbp.cs.y.m <- coxph(Surv(I9_HYPTENS_AGE, HT_EARLY) 
                       ~ SBP_SCALED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_m)

#Old
cx.sbp.cs.o.m <- coxph(Surv(I9_HYPTENS_AGE, HT_LATE) 
                       ~ SBP_SCALED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_m)

```




**Hazard ratios**


```{r HRs continuous by age}

my_hr_table_cont2(list(cx.sbp.cs.y.f, cx.sbp.cs.o.f, cx.sbp.cs.y.m,  cx.sbp.cs.o.m), 
                  c("Early female", "Late female", "Early male", "Late male")) %>%
  kable()%>%
  kable_styling(full_width =  F)


```



**Interaction term**

We will check if differences between genders is statistically different by examing the interaction term.


```{r interactions continuous score}

#All ages
cx.sbp.cs.i <- coxph(Surv(I9_HYPTENS_AGE, I9_HYPTENS) ~ SBP_SCALED*SEX_IMPUTED
                  + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)

#Early
cx.sbp.cs.y.i <- coxph(Surv(I9_HYPTENS_AGE, HT_EARLY) ~ SBP_SCALED*SEX_IMPUTED
                  + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)

#Late
cx.sbp.cs.o.i <- coxph(Surv(I9_HYPTENS_AGE, HT_LATE) ~ SBP_SCALED*SEX_IMPUTED
                  + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)


rbind(my_extr_coef(cx.sbp.cs.i, "All", ":SEX"), 
        my_extr_coef(cx.sbp.cs.y.i, "Early", ":SEX"),
        my_extr_coef(cx.sbp.cs.o.i, "Late", ":SEX")) %>%
  my_tidy_table() %>%
  kable(full_width =  F) %>%
  kable_styling()

```
  
Interaction terms are statistically significant.




## Cox model: covariate adjusted, categorized score

We will run divide data by gender and by age


**Running models**

*Female*

```{r model categorized score female}

#All ages
cx.sbp.f <- coxph(Surv(I9_HYPTENS_AGE, I9_HYPTENS) ~ SBP_CAT 
                  + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_fr)

#Young
cx.sbp.y.f <- coxph(Surv(I9_HYPTENS_AGE, HT_EARLY) ~ SBP_CAT 
                    + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_fr)

#Old
cx.sbp.o.f <- coxph(Surv(I9_HYPTENS_AGE, HT_LATE) ~ SBP_CAT 
                    + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_fr)

```

*Male*

```{r model categorized score male}
#All ages
cx.sbp.m <- coxph(Surv(I9_HYPTENS_AGE, I9_HYPTENS) ~ SBP_CAT 
                  + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_mr)

#Young
cx.sbp.y.m <- coxph(Surv(I9_HYPTENS_AGE, HT_EARLY) ~ SBP_CAT 
                    + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_mr)

#Old
cx.sbp.o.m <- coxph(Surv(I9_HYPTENS_AGE, HT_LATE) ~ SBP_CAT 
                    + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_mr)

```
  

**Interaction term**

We will check if differences between genders is statistically different by examing the interaction term.


```{r interaction categorized score}

#All ages
cx.sbp.i <- coxph(Surv(I9_HYPTENS_AGE, I9_HYPTENS) ~ SBP_CAT*SEX_IMPUTED
                  + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)

#All ages
cx.sbp.y.i <- coxph(Surv(I9_HYPTENS_AGE, HT_EARLY) ~ SBP_CAT*SEX_IMPUTED
                  + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)

#All ages
cx.sbp.o.i <- coxph(Surv(I9_HYPTENS_AGE, HT_LATE) ~ SBP_CAT*SEX_IMPUTED
                  + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)


int_table <- 
  my_extr_coef(cx.sbp.i, "", ":SEX") %>%
  my_tidy_table() %>%
  select(-name) %>%
  kable() %>%
  kable_styling(full_width =  F)

```
  
Interaction terms are statistically significant.


**Table: Hazard ratios:**


```{r HRs categorized score}

table.hr.f <- 
  my_hr_table3_notitle(cx.sbp.f, cx.sbp.y.f, cx.sbp.o.f, dfs=list(df_f, df_f, df_f)) %>%
  mutate("P-value" = sub("e(-\\d+)", "x10<sup>\\1</sup>", `P-value`, perl=T))   #html-version/proper exponents
  #mutate("P-value" = sub("e(-\\d+)", "%*%10^{\\1}", `P-value`, perl=T))        #latex-version/proper exponents

table.hr.m <- 
  my_hr_table3_notitle(cx.sbp.m, cx.sbp.y.m, cx.sbp.o.m, dfs=list(df_m, df_m, df_m)) %>% 
  select(-PRS) %>%
  mutate("P-value" = sub("e(-\\d+)", "x10<sup>\\1</sup>", `P-value`, perl=T))   #html-version

table.i.term <- 
  my_hr_table3_notitle(cx.sbp.i, cx.sbp.y.i, cx.sbp.o.i, dfs=list(df, df, df), select=":SEX") %>% 
  select(`P-value`) %>%
  mutate("P-value" = sub("e(-\\d+)", "x10<sup>\\1</sup>", `P-value`, perl=T))   #html-version

hr_table <- cbind(table.hr.f, table.hr.m, table.i.term) %>%
  kable(format="html", escape = F) %>%
  pack_rows("Early", 6, 10, bold=T) %>%
  pack_rows("Late", 11, 15, bold=T) %>%
  kable_classic_2() %>%
  add_header_above(c(" ", "Women" = 3, "Men" = 3, "Interaction" = 1), bold=T) %>%
  row_spec(0,bold=T) 

hr_table %>% save_kable("hr_table_sex.html")
hr_table

```


<details><summary>**All coefficients**</summary>


```{r all coefs categorized score}

summary(cx.sbp.f)
summary(cx.sbp.m)

summary(cx.sbp.y.f)
summary(cx.sbp.y.m)

summary(cx.sbp.o.f)
summary(cx.sbp.o.m)


```


</details>
<br>


**Selecting covariates for plot**

We used **mean** values for all numeric variables. For 'batch', we use the batch that has low significance for both data sets. AxiomGT1_b07_V2P2.calls

<details><summary>**All coefficients**</summary>

```{r select coefs for plot}
summary(cx.sbp.f)$coefficients %>%
  as.data.frame() %>%
  mutate(name = row.names(.)) %>%
  rename(pval="Pr(>|z|)") %>%
  select("name", "coef", "exp(coef)","se(coef)", "pval") %>% 
  arrange(pval)

summary(cx.sbp.m)$coefficients %>%
  as.data.frame() %>%
  mutate(name = row.names(.)) %>%
  rename(pval="Pr(>|z|)") %>%
  select("name", "coef", "exp(coef)","se(coef)", "pval") %>%
  arrange(pval)

#Low significance for both datasets: AxiomGT1_b07_V2P2.calls
 
```

</details>
<br>

Low significance for both datasets: AxiomGT1_b07_V2P2.calls


**Cox-plot adjusted, categorized**

Creating covariate combinations for the plot.

Female

```{r fit cox plot female}

df.sbp.f <- my_expand(cx.sbp.f, "SBP_CAT", df_f, batch = "AxiomGT1_b07_V2P2.calls")
fit.sbp.f <- survfit(cx.sbp.f, newdata = df.sbp.f)

```

Male

```{r fit cox plot male}

df.sbp.m <- my_expand(cx.sbp.m, "SBP_CAT", df_m, batch = "AxiomGT1_b07_V2P2.calls")
fit.sbp.m <- survfit(cx.sbp.m, newdata = df.sbp.m)

```



And the plot:

```{r cox plot, fig.width=10, fig.height=4}

#Female
p1 <- my_ggcoxplot(fit.sbp.f, df.sbp.f, "Women", legend="none")

#Male
p2 <- my_ggcoxplot(fit.sbp.m, df.sbp.m, "Men", legend="none", ylab="")+
theme(legend.position = c(0.2, 0.55))

splots <- grid.arrange(p1, p2, nrow = 1, widths = c(3,3), top=textGrob("\n", x = 0.16, y=0.2, gp=gpar(fontsize=12)))
ggsave(file = "cx_ht_prs_sex.png", plot = splots, height = 4, width = 8, dpi = 150)
ggsave(file = "cx_ht_prs_sex.eps", device = cairo_ps, plot = splots, height = 4, width = 8, dpi = 150)

#Purpose of top-parameter is to get image size same as in Figure 2; the title makes figure slightly smaller.

#lgnd <-  my_ggcoxplot(fit.sbp.f, df.sbp.f, "", legend="right") %>%
#         get_legend()

#splots <- grid.arrange(p1, p2, lgnd, nrow = 1, widths = c(3,3,1))

#ggsave(file = "cx_ht_prs_sex.png", plot = splots, height = 4, width = 10, dpi = 150)
#ggsave(file = "cx_ht_prs.pdf", plot = splots, height = 4, width = 10, dpi = 150)
#ggsave(file = "cx_ht_prs_sex.eps", device = cairo_ps, plot = splots, height = 4, width = 10, dpi = 150)

```



**Cox-plots by age**

Creating covariate combinations for the plot.

Young

```{r fit cox plot female}

df.sbp.y.f <- my_expand(cx.sbp.y.f, "SBP_CAT", df_f, batch = "AxiomGT1_b07_V2P2.calls")
fit.sbp.y.f <- survfit(cx.sbp.y.f, newdata = df.sbp.y.f)

df.sbp.y.m <- my_expand(cx.sbp.y.m, "SBP_CAT", df_m, batch = "AxiomGT1_b07_V2P2.calls")
fit.sbp.y.m <- survfit(cx.sbp.y.m, newdata = df.sbp.y.m)


```

Old

```{r fit cox plot male}

df.sbp.o.f <- my_expand(cx.sbp.o.f, "SBP_CAT", df_f, batch = "AxiomGT1_b07_V2P2.calls")
fit.sbp.o.f <- survfit(cx.sbp.o.f, newdata = df.sbp.o.f)

df.sbp.o.m <- my_expand(cx.sbp.o.m, "SBP_CAT", df_m, batch = "AxiomGT1_b07_V2P2.calls")
fit.sbp.o.m <- survfit(cx.sbp.o.m, newdata = df.sbp.o.m)

```



Plot by age

```{r cox plot by age, fig.width=10, fig.height=10}

#Young
p1 <- my_ggcoxplot(fit.sbp.y.f, df.sbp.y.f, "Women", legend="none", xlim=c(0,60), ylim=c(0,0.4))
p2 <- my_ggcoxplot(fit.sbp.y.m, df.sbp.y.m, "Men", legend="none", xlim=c(0,60), ylim=c(0,0.4), ylab="")+
theme(legend.position = c(0.2, 0.55))
splots.y <- grid.arrange(p1, p2, nrow = 1, widths = c(3,3), top=textGrob("\nA. Early-Onset Hypertension", x = 0.16, y=0.2, gp=gpar(fontsize=12)))


#Old
p1 <- my_ggcoxplot(fit.sbp.o.f, df.sbp.o.f, "Women", legend="none", xlim=c(50,80), ylim=c(0,0.8), break.time=10)
p2 <- my_ggcoxplot(fit.sbp.o.m, df.sbp.o.m, "Men", legend="none", xlim=c(50,80), ylim=c(0,0.8), break.time=10, ylab="")
splots.o <- grid.arrange(p1, p2, nrow = 1, widths = c(3,3), top=textGrob("\nB. Late-Onset Hypertension", x = 0.16, y=0.2, gp=gpar(fontsize=12)))

#Combine young & old
splots.age <- grid.arrange(splots.y, splots.o, nrow=2)
ggsave(file = "cx_ht_prs_sex_age.png", plot = splots.age, height = 8, width = 8, dpi = 150)
ggsave(file = "cx_ht_prs_sex_age.eps", device = cairo_ps, plot = splots.age, height = 8, width = 8, dpi = 150)

#ggsave(file = "cx_ht_prs_sex_y.png", plot = splots.y, height = 4, width = 8, dpi = 150)
#ggsave(file = "cx_ht_prs_sex.eps", device = cairo_ps, plot = splots, height = 4, width = 10, dpi = 150)

```


## Restricted mean survival times

Pairwise statistics between reference risk (20-80% quantile) and each other quantile is are calculated. For this purpose we need separate dataframe for each statistic of interest. 


**Creating dataframes for each cases:**


```{r rmst dfs}

q <- c("<2.5%", "2.5-20%", "80-97.5%", ">97.5%" )
df.list.sbp.f  <-  lapply(q, my_extr_cats2, cat="SBP_CAT", df=df_f)
df.list.sbp.m  <-  lapply(q, my_extr_cats2, cat="SBP_CAT", df=df_m)

```


**rmst2 is runned for all cases**

Unadjusted analysis - with covariates R is slow and output of this package does not include mean ages for adjusted.

```{r rmst model}

#Unadjusted, this works:

rmst0.sbp.f <- lapply(df.list.sbp.f, function(x) rmst2(x$I9_HYPTENS_AGE, x$I9_HYPTENS, x$CAT, tau=95))
rmst0.sbp.m <- lapply(df.list.sbp.m, function(x) rmst2(x$I9_HYPTENS_AGE, x$I9_HYPTENS, x$CAT, tau=95))

```


**Results are extracted for the table**

SBP for females and males

```{r rmst tables}

#Results extracted and collected to table

rmst.table <- 
  my_rmst_coefs(rmst0.sbp.f) %>%
  bind_rows(my_rmst_coefs(rmst0.sbp.m)) %>%
  my_tidy_table("0", 1) %>%
  add_row(q="Women", est="", pval="", lower_mean=NA, upper_mean=NA, mean=NA,  .before = 1) %>%
  add_row(q="", est="", pval="", lower_mean=NA, upper_mean=NA, mean=NA,  .before = 7) %>%
  add_row(q="Men", est="", pval="", lower_mean=NA, upper_mean=NA, mean=NA, .before = 8)


#rmst.table divided to two and formatted for forestplot

rmst.means <- rmst.table %>%
  select(mean, lower_mean, upper_mean) %>%
  rename(lower = lower_mean, upper=upper_mean) %>%
  add_row(mean=NA, lower=NA, upper=NA, .before = 1)

rmst.text <- rmst.table %>%
 select(q, est, pval) %>%
  mutate(pval = sub("e(-\\d+)", "%*%10^{\\1}", pval, perl=T)) %>%     #to create proper exponents...
  add_row(q=NA, est="Difference, Years (95% CI)", pval="P-value", .before = 1)

#To create proper exponents... I need actual values as 'expression' and other lines as 'character'.
#list format is needed to accomodate different datatypes; parse can convert character unlike 'as.expression'
pvals <- lapply(rmst.text$pval, function(x) {ifelse(grepl("10",x), parse(text = x), x)})
rmst.text.list <- list(as.list(rmst.text$q), as.list(rmst.text$est), pvals)     
```



**Forestplots**

```{r rmst forestplot,fig.width=9, fig.height=5.5}

setEPS()
postscript(file = 'rmst_unadj_bysex.eps', width = 9, height = 4.5) 

my_rmst_plot(rmst.text.list, rmst.means, 
             clip=c(65,90), xticks = c(65, 70, 75,  80, 85), zero=65)
dev.off()

my_rmst_plot(rmst.text.list, rmst.means, 
             clip=c(65,90), xticks = c(65, 70, 75,  80, 85), zero=65)

```


<details><summary>**Test: Only time differences shown**</summary>

```{r rmst difference only model}
#Let's test this too
rmst.diffs <- 
  my_rmst_coefs(rmst0.sbp.f) %>%
  bind_rows(my_rmst_coefs(rmst0.sbp.m)) %>%
  select(est, lower, upper) %>%
  mutate_all(replace_na, 0) %>%
  add_row(est=NA, lower=NA, upper=NA, .before = 1) %>%
  add_row(est=NA, lower=NA, upper=NA, .before = 1) %>%
  add_row(est=NA, lower=NA, upper=NA, .before = 8) %>%
  add_row(est=NA, lower=NA, upper=NA, .before = 9) %>%
  rename(mean=est) %>%
  mutate_all(as.numeric)
 
```



And plot...


```{r rmst difference only plot,fig.width=9, fig.height=5.5}

setEPS()
postscript(file = 'rmst_diffs_unadj_bysex.eps', width = 9, height = 4.5) 

my_rmst_plot(rmst.text.list, rmst.diffs, 
             clip=c(-11, 10), xticks = c(-10, -5, 0, 5, 10), zero=0,
             xlab = "Age difference, years")

dev.off()

my_rmst_plot(rmst.text.list, rmst.diffs, 
             clip=c(-11, 10), xticks = c(-10, -5, 0, 5, 10), zero=0,
             xlab = "Age difference, years")


```


</details>
<br>



# BP PRSs and cardiovascular outcomes


## Proportional hazard assumption: log log curve:


**Model by survfit**


```{r}

#CHD
km.sbp.chd.f <- survfit(Surv(I9_CHD_AGE, I9_CHD) ~ SBP_CAT, data=df_f)
km.sbp.chd.m <- survfit(Surv(I9_CHD_AGE, I9_CHD) ~ SBP_CAT, data=df_m)

```



**log log curve:**


```{r, fig.width=10, fig.height=8}

cols= c("skyblue", "blue", "black", "red", "salmon")
labels= c("<2.5%","2.5-20%","20-80%","80-97.5%",">97.5%")

par(mfrow=c(1,2))

plot(km.sbp.chd.f, fun="cloglog", xlim=c(40,100), main="SBP CHD female: log log", xlab="Age", col = cols)
plot(km.sbp.chd.m, fun="cloglog", xlim=c(40,100), main="SBP CHD male: log log", xlab="Age", col = cols)

par(mfrow=c(1,1))

```


## Cox model: covariate adjusted, continuous score


**Running models**


*Female*

```{r female continuous by age}

#All
cx.sbp.chd.cs.f <- coxph(Surv(I9_CHD_AGE, I9_CHD) ~ SBP_SCALED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_f)

#Young
cx.sbp.chd.cs.y.f <- coxph(Surv(I9_CHD_AGE, CHD_EARLY) 
                       ~ SBP_SCALED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_f)

#Old
cx.sbp.chd.cs.o.f <- coxph(Surv(I9_CHD_AGE, CHD_LATE) 
                       ~ SBP_SCALED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_f)

```

*Male*

```{r male continuous by age}

#All
cx.sbp.chd.cs.m <- coxph(Surv(I9_CHD_AGE, I9_CHD) ~ SBP_SCALED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_m)

#Young
cx.sbp.chd.cs.y.m <- coxph(Surv(I9_CHD_AGE, CHD_EARLY) 
                       ~ SBP_SCALED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_m)

#Old
cx.sbp.chd.cs.o.m <- coxph(Surv(I9_CHD_AGE, CHD_LATE) 
                       ~ SBP_SCALED + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_m)

```

**Hazard ratios**


```{r HRs continuous by age}

my_hr_table_cont2(list(cx.sbp.chd.cs.f, cx.sbp.chd.cs.m, cx.sbp.chd.cs.y.f, cx.sbp.chd.cs.o.f, cx.sbp.chd.cs.y.m,  cx.sbp.chd.cs.o.m), 
                  c("CHD female", "CHD male","CHD early female", "CHD late female", "CHD early male", "CHD late male")) %>%
  kable()%>%
  kable_styling(full_width =  F)


```


<br>

<details><summary>**All coefficients**</summary>


```{r}


summary(cx.sbp.chd.cs.f)
summary(cx.sbp.chd.cs.m)

summary(cx.sbp.chd.cs.y.f)
summary(cx.sbp.chd.cs.y.m)
summary(cx.sbp.chd.cs.o.f)
summary(cx.sbp.chd.cs.o.m)


```


</details>
<br>



**Interaction term**

We will check if differences between genders is statistically different by examing the interaction term.


```{r interactions continuous score}

#All ages
cx.sbp.chd.cs.i <- coxph(Surv(I9_CHD_AGE, I9_CHD) ~ SBP_SCALED*SEX_IMPUTED
                  + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)

#Early
cx.sbp.chd.cs.y.i <- coxph(Surv(I9_CHD_AGE, CHD_EARLY) ~ SBP_SCALED*SEX_IMPUTED
                  + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)

#Late
cx.sbp.chd.cs.o.i <- coxph(Surv(I9_CHD_AGE, CHD_LATE) ~ SBP_SCALED*SEX_IMPUTED
                  + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)


rbind(my_extr_coef(cx.sbp.chd.cs.i, "All", ":SEX"), 
        my_extr_coef(cx.sbp.chd.cs.y.i, "Early", ":SEX"),
        my_extr_coef(cx.sbp.chd.cs.o.i, "Late", ":SEX")) %>%
  my_tidy_table() %>%
  kable(full_width =  F) %>%
  kable_styling()

```
  


## Cox model: covariate adjusted, categorized score



## Cox model for CHD: covariate adjusted, categorized score

We will run divide data by gender and by age


**Running models**

*Female*

```{r model categorized score female}

#All ages
cx.sbp.chd.f <- coxph(Surv(I9_CHD_AGE, I9_CHD) ~ SBP_CAT 
                  + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_fr)

#Young
cx.sbp.chd.y.f <- coxph(Surv(I9_CHD_AGE, CHD_EARLY) ~ SBP_CAT 
                    + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_fr)

#Old
cx.sbp.chd.o.f <- coxph(Surv(I9_CHD_AGE, CHD_LATE) ~ SBP_CAT 
                    + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_fr)

```

*Male*

```{r model categorized score male}
#All ages
cx.sbp.chd.m <- coxph(Surv(I9_CHD_AGE, I9_CHD) ~ SBP_CAT 
                  + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_mr)

#Young
cx.sbp.chd.y.m <- coxph(Surv(I9_CHD_AGE, CHD_EARLY) ~ SBP_CAT 
                    + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_mr)

#Old
cx.sbp.chd.o.m <- coxph(Surv(I9_CHD_AGE, CHD_LATE) ~ SBP_CAT 
                    + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_mr)

```
  

**Interaction term**

We will check if differences between genders is statistically different by examing the interaction term.


```{r interaction categorized score}

#All ages
cx.sbp.chd.i <- coxph(Surv(I9_CHD_AGE, I9_CHD) ~ SBP_CAT*SEX_IMPUTED
                  + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)

#All ages
cx.sbp.chd.y.i <- coxph(Surv(I9_CHD_AGE, CHD_EARLY) ~ SBP_CAT*SEX_IMPUTED
                  + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)

#All ages
cx.sbp.chd.o.i <- coxph(Surv(I9_CHD_AGE, CHD_LATE) ~ SBP_CAT*SEX_IMPUTED
                  + batch + BL_YEAR + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data=df_r)


int_table.chd <- 
  my_extr_coef(cx.sbp.chd.i, "", ":SEX") %>%
  my_tidy_table() %>%
  select(-name) %>%
  kable() %>%
  kable_styling(full_width =  F)

```


**Table: Hazard ratios:**


```{r HRs categorized score}

table.hr.chd.f <- 
  my_hr_table3_notitle(cx.sbp.chd.f, cx.sbp.chd.y.f, cx.sbp.chd.o.f, dfs=list(df_f, df_f, df_f)) %>%
  mutate("P-value" = sub("e(-\\d+)", "x10<sup>\\1</sup>", `P-value`, perl=T))   #html-version/proper exponents
  #mutate("P-value" = sub("e(-\\d+)", "%*%10^{\\1}", `P-value`, perl=T))        #latex-version/proper exponents

table.hr.chd.m <- 
  my_hr_table3_notitle(cx.sbp.chd.m, cx.sbp.chd.y.m, cx.sbp.chd.o.m, dfs=list(df_m, df_m, df_m)) %>% 
  select(-PRS) %>%
  mutate("P-value" = sub("e(-\\d+)", "x10<sup>\\1</sup>", `P-value`, perl=T))   #html-version

table.i.chd.term <- 
  my_hr_table3_notitle(cx.sbp.chd.i, cx.sbp.chd.y.i, cx.sbp.chd.o.i, dfs=list(df, df, df), select=":SEX") %>% 
  select(`P-value`) %>%
  mutate("P-value" = sub("e(-\\d+)", "x10<sup>\\1</sup>", `P-value`, perl=T))   #html-version

hr_table.chd <- cbind(table.hr.chd.f, table.hr.chd.m, table.i.chd.term) %>%
  kable(format="html", escape = F) %>%
  pack_rows("Early", 6, 10, bold=T) %>%
  pack_rows("Late", 11, 15, bold=T) %>%
  kable_classic_2() %>%
  add_header_above(c(" ", "Women" = 3, "Men" = 3, "Interaction" = 1), bold=T) %>%
  row_spec(0,bold=T) 

hr_table.chd %>% save_kable("hr_table_sbpprs_chd_sex.html")
hr_table.chd

```


<details><summary>**All coefficients**</summary>


```{r all coefs categorized score}

summary(cx.sbp.chd.f)
summary(cx.sbp.chd.m)

summary(cx.sbp.chd.y.f)
summary(cx.sbp.chd.y.m)

summary(cx.sbp.chd.o.f)
summary(cx.sbp.chd.o.m)


```


</details>
<br>















